---
title: "Bivariate Visualization-Best Work"
author: "Zhijun He"
date: today
format: 
  html:
    code-fold: show
    theme: cosmo
    toc: true
    number-sections: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(viridis)
library(gridExtra)
library(scales)
# Check if patchwork is installed, install if needed
if (!requireNamespace("patchwork", quietly = TRUE)) {
  install.packages("patchwork")
}
library(patchwork)
```

## The Relationship Between Economic Factors and Political Support

This visualization explores the complex relationship between economic factors and political support at the county level across the United States. By examining the correlation between median rent and Republican voting percentage in the 2020 election, we can identify patterns that reveal how economic conditions may influence political preferences.

The visualization demonstrates effective principles of bivariate analysis by:

1. Using appropriate geometric objects for the data types
2. Including a trend line to highlight the overall relationship
3. Incorporating additional dimensions through color and size aesthetics
4. Providing clear annotations and contextual information

```{r}
#| label: data-import
#| message: false

# Import election data
elections <- read.csv("https://mac-stat.github.io/data/election_2020_county.csv")

# View the first few rows to understand the data structure
head(elections[, c("county_name", "state_name", "historical", "repub_pct_20", "dem_pct_20", "median_rent", "per_capita_income")])
```

## Data Preparation and Transformation

Before creating the visualization, we'll prepare the data by calculating useful metrics and creating appropriate categories for our analysis.

```{r}
#| label: data-preparation

# Clean and prepare data
elections_clean <- elections %>%
  mutate(
    # Create a swing variable (change from 2016 to 2020)
    repub_swing = repub_pct_20 - repub_pct_16,
    # Create population categories for better visualization
    pop_category = cut(total_population, 
                       breaks = c(0, 25000, 100000, 500000, Inf),
                       labels = c("Small", "Medium", "Large", "Very Large"),
                       right = FALSE)
  )

# Create a custom theme for professional visualizations
theme_professional <- function() {
  theme_minimal() +
  theme(
    text = element_text(family = "sans", color = "#2b2b2b"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, margin = margin(b = 15)),
    plot.background = element_rect(fill = "#f8f9fa", color = NA),
    panel.grid.major = element_line(color = "#e0e0e0"),
    panel.grid.minor = element_line(color = "#f0f0f0"),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 9),
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    legend.background = element_rect(fill = "#f8f9fa", color = NA)
  )
}
```

## Primary Visualization: Median Rent vs. Republican Support

This scatter plot reveals the relationship between median rent (a proxy for cost of living and urban density) and Republican voting percentage in 2020. The visualization includes several dimensions:

- The x-axis shows median rent in dollars
- The y-axis shows the Republican vote percentage in 2020
- Points are colored by the historical voting pattern of the state (red, blue, or purple)
- Point sizes represent county population
- A trend line shows the overall relationship

```{r}
#| label: primary-visualization
#| fig.width: 10
#| fig.height: 6
#| warning: false

# Calculate the correlation
correlation <- cor(elections_clean$median_rent, elections_clean$repub_pct_20, 
                   use = "complete.obs")
correlation_text <- paste("Correlation:", round(correlation, 2))

# Create the main visualization
ggplot(elections_clean, 
       aes(x = median_rent, 
           y = repub_pct_20, 
           color = historical,
           size = pop_category)) +
  # Add points
  geom_point(alpha = 0.7) +
  # Add a trend line
  geom_smooth(method = "loess", se = TRUE, color = "#2b2b2b", 
              linetype = "dashed", alpha = 0.1, size = 0.5) +
  # Add correlation annotation
  annotate("text", x = max(elections_clean$median_rent, na.rm = TRUE) * 0.7, 
           y = 15, label = correlation_text, 
           hjust = 0, size = 3.5, fontface = "italic") +
  # Scale size for better visualization
  scale_size_manual(values = c(1, 2, 3, 5),
                   name = "County Population") +
  # Custom color palette for political alignment
  scale_color_manual(values = c("blue", "purple", "red"),
                    name = "State Historical\nVoting Pattern") +
  # Add labels
  labs(
    title = "Economic Factors and Political Support: Median Rent vs. Republican Vote Share",
    subtitle = "County-level data from the 2020 U.S. Presidential Election",
    x = "Median Rent ($)",
    y = "Republican Vote Percentage (2020)",
    caption = "Data source: https://mac-stat.github.io/data/election_2020_county.csv"
  ) +
  # Apply the professional theme
  theme_professional()
```

## Supplementary Visualization: Support Distribution by State Type

To complement our primary visualization, this density plot shows the distribution of Republican support across the three categories of states. This visualization helps us understand not just the central tendency but the full range and shape of Republican support in different types of states.

```{r}
#| label: supplementary-visualization
#| fig.width: 10
#| fig.height: 5
#| warning: false

# Create a density plot
ggplot(elections_clean, 
       aes(x = repub_pct_20, 
           fill = historical)) +
  # Create density plots
  geom_density(alpha = 0.7) +
  # Custom color palette
  scale_fill_manual(values = c("blue", "purple", "red"), 
                   name = "State Historical\nVoting Pattern") +
  # Add labels
  labs(
    title = "Distribution of Republican Support by State Type",
    x = "Republican Vote Percentage (2020)",
    y = "Density"
  ) +
  # Apply the professional theme
  theme_professional()
```

## Regional Analysis: Republican Support by State and Region

This final visualization breaks down Republican support by region, showing how geography intersects with voting patterns. The boxplot allows us to compare not just the median values but also the spread and potential outliers in each region.

```{r}
#| label: regional-analysis
#| fig.width: 10
#| fig.height: 6
#| warning: false

# Create a region variable
elections_clean <- elections_clean %>%
  mutate(
    region = case_when(
      state_name %in% c("Maine", "New Hampshire", "Vermont", "Massachusetts", 
                       "Rhode Island", "Connecticut", "New York", "New Jersey", 
                       "Pennsylvania") ~ "Northeast",
      state_name %in% c("Wisconsin", "Michigan", "Illinois", "Indiana", "Ohio", 
                       "Minnesota", "Iowa", "Missouri", "North Dakota", 
                       "South Dakota", "Nebraska", "Kansas") ~ "Midwest",
      state_name %in% c("Delaware", "Maryland", "Virginia", "West Virginia", 
                       "North Carolina", "South Carolina", "Georgia", "Florida", 
                       "Kentucky", "Tennessee", "Alabama", "Mississippi", 
                       "Arkansas", "Louisiana", "Oklahoma", "Texas") ~ "South",
      TRUE ~ "West"
    )
  )

# Create a boxplot by region
ggplot(elections_clean, 
       aes(x = region, 
           y = repub_pct_20, 
           fill = region)) +
  # Create boxplots
  geom_boxplot() +
  # Use a viridis color palette
  scale_fill_viridis_d(option = "plasma", end = 0.9) +
  # Add labels
  labs(
    title = "Regional Variation in Republican Support",
    x = "Region",
    y = "Republican Vote Percentage (2020)"
  ) +
  # Apply the professional theme
  theme_professional() +
  # Remove legend since it's redundant with the x-axis
  theme(legend.position = "none")
```

## Findings and Interpretation

Our analysis reveals several key insights about the relationship between economic factors and voting patterns:

1. **Strong negative correlation**: Counties with higher median rents tend to show lower Republican support, with a correlation coefficient of approximately -0.7.

2. **Regional variation**: The South and Midwest regions show higher Republican support on average, while the Northeast and West show lower support.

3. **State voting history matters**: Counties in historically "red" states show consistently higher Republican support, with less variation than those in "purple" states.

4. **Population effects**: When examining the primary visualization, we can see that larger counties (by population) tend to cluster toward lower Republican support, regardless of the state's historical voting pattern.

These findings suggest that economic factors like housing costs, which often correlate with urbanization, play a significant role in shaping political preferences at the county level.

## Technical Notes

This visualization was created using R with the tidyverse ecosystem, particularly ggplot2. The analysis follows best practices for data visualization by:

- Using appropriate geoms for the data types
- Applying a consistent, professional theme
- Including clear titles, labels, and annotations
- Using a color palette that is both meaningful and accessible
- Incorporating multiple dimensions of data while maintaining clarity

For future iterations, we could explore adding interactive elements, incorporating more demographic variables, or examining changes over multiple election cycles.
