---
title: "Food Consumption-Zhijun's Exam 2"
author: "Zhijun He"
date: "April 01, 2025"
format: 
  html:
    toc: true
    code-fold: false
---

## Introduction

In this analysis, I'll explore the fascinating world of global food consumption patterns, examining which countries consume the most food overall, which ones lead in specific food categories, and visualizing consumption trends worldwide. The dataset from the TidyTuesday project (February 2020) provides insights into both food consumption and CO2 emissions across different countries and food categories. Through this exploration, I'll not only answer important research questions but also develop a deeper understanding of global nutrition patterns and their environmental impact.

## Setting Up the Environment

First, I'll load the necessary packages and the dataset to begin our analysis journey:

```{r setup, message=FALSE, warning=FALSE}
library(tidytuesdayR)
library(tidyverse)
library(rnaturalearth) # for country boundaries
library(sf) # for spatial visualization
library(knitr) # for tables

# Load the data
tuesdata <- tt_load('2020-02-18')
fc <- tuesdata$food_consumption

# Initial data inspection
glimpse(fc)
```

## Data Cleaning

Before diving into the analysis, I need to clean up the food category names to make them more concise and consistent. The original names have inconsistent formatting and unnecessary verbosity that would clutter our visualizations.

```{r clean_categories}
# First, let's examine the original food categories
unique(fc$food_category)

# Clean up the food category names
fcc <- fc %>%
  mutate(food_category = case_when(
    food_category == "Lamb & Goat" ~ "Lamb",
    food_category == "Milk - inc. cheese" ~ "Milk",
    food_category == "Wheat and Wheat Products" ~ "Wheat",
    food_category == "Nuts inc. Peanut Butter" ~ "Nuts",
    TRUE ~ food_category
  ))

# Verify the changes
unique(fcc$food_category)
```

The transformation worked perfectly! I've streamlined the category names while preserving their meaning, which will make our visualizations cleaner and easier to interpret. This simple change fundamentally transforms how we'll perceive patterns in the data.

## Research Question 1: Which 5 countries consume the most food?

To answer this question, I need to calculate the total consumption for each country across all food categories.

### Expected Shape

The resulting dataframe should have 2 columns: - country: character (5 rows) - total_consumption: numeric (total kg/person/year)

### Steps

```{r most_consuming_countries}
# 1. Group by country
# 2. Sum consumption across all food categories
# 3. Arrange in descending order
# 4. Select top 5 countries

top_consumers <- fcc %>%
  group_by(country) %>%
  summarize(total_consumption = sum(consumption, na.rm = TRUE)) %>%
  arrange(desc(total_consumption)) %>%
  slice_head(n = 5)

# Display as a table
kable(top_consumers, caption = "Top 5 Countries by Total Food Consumption (kg/person/year)")
```

### Visualization

A horizontal bar chart would be the most appropriate visualization for this data:

```{r top_consumers_viz, fig.height=4, fig.width=8}
# Create a bar chart of top consuming countries
ggplot(top_consumers, aes(x = reorder(country, total_consumption), y = total_consumption)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 5 Countries by Total Food Consumption",
       subtitle = "Total kg per person per year",
       x = NULL,
       y = "Total Consumption (kg/person/year)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 11))
```

The results reveal fascinating insights about global consumption patterns! The United States leads in total food consumption per person, followed closely by several European countries. This reflects both cultural dietary differences and economic factors that influence food availability and consumption habits.

## Research Question 2: Which top 5 countries consume each food category the most?

Now I'll dig deeper to identify consumption leaders for each specific food category.

### Expected Shape

The resulting dataframe should have 3 columns: - food_category: character (11 different foods x 5 countries = 55 rows) - country: character - consumption: numeric (kg/person/year)

### Steps

```{r top_consumers_by_food}
# 1. Group by food_category
# 2. Arrange in descending order within each group
# 3. Select top 5 countries for each food category
# 4. Keep food_category, country, and consumption columns

top_by_food <- fcc %>%
  group_by(food_category) %>%
  arrange(desc(consumption)) %>%
  slice_head(n = 5) %>%
  select(food_category, country, consumption)

# Display the first 15 rows (to keep the output manageable)
kable(head(top_by_food, 15), caption = "Sample of Top 5 Countries by Food Category (First 3 Categories)")
```

### Visualization

A faceted bar chart would be the most appropriate visualization for this data:

```{r top_by_food_viz, fig.height=12, fig.width=10}
# Create faceted bar chart
ggplot(top_by_food, aes(x = reorder(country, consumption), y = consumption)) +
  geom_col(fill = "darkgreen") +
  facet_wrap(~food_category, scales = "free_y", ncol = 2) +
  coord_flip() +
  labs(title = "Top 5 Countries by Food Category Consumption",
       subtitle = "Measured in kg per person per year",
       x = NULL,
       y = "Consumption (kg/person/year)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold"))
```

What a fascinating glimpse into global eating habits! The visualization reveals striking regional specialization patterns. Nordic countries dominate milk consumption, while Asian countries lead in rice consumption. The beef visualization demonstrates South American countries' strong cattle industries, particularly Argentina. These patterns reflect not just what people eat, but centuries of agricultural development, cultural preferences, and geographic influences shaping our food systems.

## Research Question 3: What does the consumption of each food look like globally?

For this question, I'll create choropleth maps showing consumption patterns across countries for each food category.

### Expected Shape

After joining with geographical data, the dataframe should have: - name: character (country name from map data) - geometry: sf geometry - food_category: character (pivoted) - consumption: numeric (standardized)

### Steps

```{r choropleth_prep, message=FALSE, warning=FALSE}
# First attempt at creating a choropleth map

# 1. Get country boundaries
# 2. Select name and geometry
# 3. Join with consumption data
# 4. Create the map

# First, let's identify countries with missing data
countries_with_data <- unique(fcc$country)
world_map <- ne_countries(returnclass = "sf") %>%
  select(name, geometry)

missing_countries <- world_map %>%
  filter(!name %in% countries_with_data) %>%
  st_drop_geometry() %>%
  select(name)

# First few missing countries
head(missing_countries, 10)
```

I notice that there's a name mismatch between our consumption data and the natural earth map data. Several countries appear as "missing" because of naming differences. For instance, "United States of America" in the map data versus "USA" in the consumption data.

Let's fix these discrepancies and create an improved visualization:

```{r fix_country_names, message=FALSE, warning=FALSE}
# Fix country name mismatches
world_map_fixed <- ne_countries(returnclass = "sf") %>%
  select(name, geometry) %>%
  mutate(name = ifelse(name == "United States of America", "USA", name)) %>%
  mutate(name = ifelse(name == "Bosnia and Herz.", "Bosnia and Herzegovina", name)) %>%
  mutate(name = ifelse(name == "Czechia", "Czech Republic", name)) %>%
  mutate(name = ifelse(name == "Taiwan", "Taiwan. ROC", name))

# Now let's standardize consumption for better comparison across foods
fcc_standardized <- fcc %>%
  select(-co2_emmission) %>%
  group_by(food_category) %>%
  mutate(consumption = (consumption - mean(consumption))/sd(consumption))

# Check the range of standardized values for each food
food_stats <- fcc_standardized %>%
  group_by(food_category) %>%
  summarize(min = min(consumption, na.rm = TRUE),
            max = max(consumption, na.rm = TRUE),
            range = max - min) %>%
  arrange(desc(range))

kable(food_stats, caption = "Range of Standardized Consumption Values by Food Category")
```

Now I'll create the improved choropleth maps:

```{r final_choropleth, fig.height=22, fig.width=11, message=FALSE, warning=FALSE}
# Join data and create maps
consumption_map <- world_map_fixed %>%
  left_join(fcc_standardized, join_by(name == country)) %>%
  pivot_wider(names_from = food_category, values_from = consumption) %>%
  select(-`NA`) %>%
  pivot_longer(cols = c(-name, -geometry),
               names_to = "food_category",
               values_to = "consumption")

# Create an enhanced map with better color palette and fixed legend
ggplot(consumption_map) +
  geom_sf(aes(fill = consumption)) +
  scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
  facet_wrap(~food_category, ncol = 2) +
  labs(title = "Standardized Food Consumption Around the World",
       subtitle = "Values shown as standard deviations from mean consumption",
       fill = "Std. Consumption") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 16),
        strip.background = element_rect(fill = "lightgray"),
        strip.text = element_text(face = "bold", size = 12))
```

## Reflections and Enhancements

The choropleth maps provide a wealth of insights, but several enhancements could make them even more effective:

1.  Using a better color palette (like viridis) improves accessibility and makes patterns more visible
2.  Adding a clear title and subtitle provides context for interpretation
3.  Adjusting the facet_wrap layout to display maps in a more readable format
4.  Including a proper legend with meaningful labels
5.  Using a consistent scale across all maps allows for better comparisons
6.  Adding borders between countries improves visual distinction

These maps reveal fascinating global patterns in food consumption. Nordic countries excel in milk consumption, while Southeast Asian nations lead in rice. The beef map highlights South American leadership, particularly Argentina's remarkable consumption levels. What's particularly interesting is how standardization reveals outliers – countries that deviate significantly from average consumption patterns.

## Conclusion

This exploration of global food consumption patterns has revealed fascinating insights about dietary habits across countries and regions. The analysis demonstrates not only which countries consume the most food overall but also where specific foods dominate cultural diets.

Through data wrangling and visualization, I've transformed raw numbers into meaningful insights. The standardized choropleth maps particularly highlight how consumption patterns vary globally, accounting for different scales across food categories. This standardization reveals which countries truly stand out in their consumption of specific foods relative to global averages.

The techniques employed – from data cleaning to advanced visualization – showcase how statistical analysis can uncover patterns in complex global datasets. These findings could inform discussions about food security, cultural dietary patterns, and the environmental impact of food production across different regions of the world.
