---
title: "Australia Fires: A Spatial and Temporal Analysis"
author: "Zhijun's Solo Project"
date: "2025-04-10"
format: 
  html:
    theme: cosmo
    toc: true
    code-fold: true
    code-tools: true
execute:
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(sf)
library(viridis)
library(scales)
library(knitr)
library(kableExtra)
library(ggrepel)
library(patchwork)
library(lubridate)
library(showtext)
library(leaflet)
library(RColorBrewer)

# Add fonts
font_add_google("Roboto", "roboto")
font_add_google("Roboto Condensed", "roboto_condensed")
showtext_auto()

# Set theme for consistent visualization
theme_set(
  theme_minimal(base_family = "roboto") +
    theme(
      plot.title = element_text(family = "roboto_condensed", face = "bold", size = 16),
      plot.subtitle = element_text(family = "roboto", size = 12),
      axis.title = element_text(family = "roboto", size = 11),
      legend.title = element_text(family = "roboto_condensed", size = 11),
      legend.text = element_text(family = "roboto", size = 10)
    )
)

# Set seed for reproducibility
set.seed(42)
```

## Introduction

Australia faced one of its most devastating bushfire seasons in 2019-2020. As someone who has always been fascinated by natural disasters and their environmental impacts, I found myself drawn to this crisis. Growing up near bushland, I witnessed smaller bushfires firsthand, but the scale of the 2019-2020 fires was unprecedented. The smoke that blanketed cities, the reddened skies, and the heartbreaking images of injured wildlife left an indelible impression on me.

In this analysis, I explore the 2019-2020 Australian bushfires through data visualization and spatial analysis. Using the TidyTuesday dataset, I examine the relationship between climate conditions and fire occurrences, focusing particularly on New South Wales (NSW), one of the hardest-hit regions.

Through this work, I hope to contribute to our understanding of how climate patterns relate to fire activity. Such insights are increasingly important as climate change continues to influence fire regimes globally. My personal connection to Australia's landscapes drives my interest in preserving these unique ecosystems for future generations.

## Data Acquisition and Preparation

For this analysis, I'm using data from the TidyTuesday project, which includes three main datasets:

1. NASA MODIS satellite fire detection data
2. Rainfall data from Australian weather stations
3. Temperature data from Australian weather stations

```{r load-data}
# Load the datasets
rainfall <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-01-07/rainfall.csv')
temperature <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-01-07/temperature.csv')
nasa_fire <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-01-07/MODIS_C6_Australia_and_New_Zealand_7d.csv')
```

Let's first look at the structure of each dataset:

```{r examine-data}
# Basic information about each dataset
rainfall_summary <- tibble(
  Dataset = "Rainfall",
  Rows = nrow(rainfall),
  Columns = ncol(rainfall),
  Time_Period = paste(min(rainfall$year, na.rm = TRUE), "-", max(rainfall$year, na.rm = TRUE)),
  Cities = paste(unique(rainfall$city_name), collapse = ", ")
)

temperature_summary <- tibble(
  Dataset = "Temperature",
  Rows = nrow(temperature),
  Columns = ncol(temperature),
  Time_Period = paste(min(year(temperature$date), na.rm = TRUE), "-", max(year(temperature$date), na.rm = TRUE)),
  Cities = paste(unique(temperature$city_name), collapse = ", ")
)

nasa_fire_summary <- tibble(
  Dataset = "NASA Fire Data",
  Rows = nrow(nasa_fire),
  Columns = ncol(nasa_fire),
  Time_Period = paste(min(nasa_fire$acq_date, na.rm = TRUE), "-", max(nasa_fire$acq_date, na.rm = TRUE)),
  Cities = "N/A (Geospatial data)"
)

bind_rows(rainfall_summary, temperature_summary, nasa_fire_summary) %>%
  kable(caption = "Dataset Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Let's clean and transform the data for analysis:

```{r clean-data}
# Prepare rainfall data
rainfall_clean <- rainfall %>%
  mutate(
    date = ymd(paste(year, month, day, sep = "-")),
    month_year = floor_date(date, "month")
  ) %>%
  # Filter to the last few years for recent trend analysis
  filter(year >= 2016)

# Prepare temperature data
temperature_clean <- temperature %>%
  mutate(month_year = floor_date(date, "month")) %>%
  # Filter to the last few years for recent trend analysis
  filter(year(date) >= 2016)

# Prepare fire data
fire_clean <- nasa_fire %>%
  mutate(
    date = as.Date(acq_date),
    # Extract Australian fires only (rough bounding box)
    is_australia = longitude >= 110 & longitude <= 155 & latitude >= -45 & latitude <= -10,
    # Convert confidence to numeric for analysis
    confidence_num = case_when(
      confidence == "low" ~ 25,
      confidence == "nominal" ~ 50,
      confidence == "high" ~ 75,
      TRUE ~ NA_real_
    ),
    # Flag night vs day
    is_night = daynight == "N"
  ) %>%
  filter(is_australia)

# Define Australia states bounding boxes (approximated)
# These are rough approximations and would be better with actual shapefiles
nsw_bbox <- list(
  lon_min = 141, lon_max = 154,
  lat_min = -37.5, lat_max = -28
)

victoria_bbox <- list(
  lon_min = 141, lon_max = 150,
  lat_min = -39, lat_max = -34
)

# Tag fires by state (approximate)
fire_clean <- fire_clean %>%
  mutate(
    in_nsw = longitude >= nsw_bbox$lon_min & longitude <= nsw_bbox$lon_max &
      latitude >= nsw_bbox$lat_min & latitude <= nsw_bbox$lat_max,
    in_victoria = longitude >= victoria_bbox$lon_min & longitude <= victoria_bbox$lon_max &
      latitude >= victoria_bbox$lat_min & latitude <= victoria_bbox$lat_max
  )
```

## Climate Analysis

Before examining the fires, let's understand the climate context during this period. Australia had been experiencing severe drought conditions leading up to the 2019-2020 fire season.

```{r rainfall-analysis}
# Analyze rainfall patterns
monthly_rainfall <- rainfall_clean %>%
  group_by(city_name, month_year) %>%
  summarize(total_rainfall = sum(rainfall, na.rm = TRUE), .groups = "drop")

# Plot rainfall trends
ggplot(monthly_rainfall, aes(x = month_year, y = total_rainfall, color = city_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_viridis_d(option = "turbo") +
  labs(
    title = "Monthly Rainfall Trends (2016-2020)",
    subtitle = "Data from major Australian cities",
    x = "Month",
    y = "Total Monthly Rainfall (mm)",
    color = "City"
  ) +
  theme(legend.position = "bottom") +
  scale_x_date(date_breaks = "4 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Now let's look at temperature patterns:

```{r temperature-analysis}
# Calculate monthly average temperatures
monthly_temp <- temperature_clean %>%
  group_by(city_name, month_year, temp_type) %>%
  summarize(avg_temp = mean(temperature, na.rm = TRUE), .groups = "drop")

# Plot temperature trends
ggplot(monthly_temp, aes(x = month_year, y = avg_temp, color = city_name, linetype = temp_type)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "inferno") +
  labs(
    title = "Monthly Temperature Trends (2016-2020)",
    subtitle = "Maximum and minimum temperatures",
    x = "Month",
    y = "Average Temperature (°C)",
    color = "City",
    linetype = "Type"
  ) +
  theme(legend.position = "bottom") +
  scale_x_date(date_breaks = "4 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The combination of low rainfall and high temperatures created dangerous fire conditions across Australia, particularly in late 2019 and early 2020.

## Fire Activity Analysis

Now, let's examine the NASA MODIS fire detection data:

```{r fire-overview}
# Summarize fire detections by date
daily_fires <- fire_clean %>%
  group_by(date) %>%
  summarize(
    fire_count = n(),
    avg_frp = mean(frp, na.rm = TRUE),
    avg_confidence = mean(confidence_num, na.rm = TRUE),
    .groups = "drop"
  )

# Plot fire activity over time
ggplot(daily_fires, aes(x = date, y = fire_count)) +
  geom_line(color = "#FF5500", linewidth = 1) +
  geom_point(aes(size = avg_frp, color = avg_confidence), alpha = 0.7) +
  scale_color_viridis_c(option = "magma") +
  scale_size_continuous(range = c(1, 5)) +
  labs(
    title = "Fire Detection Activity in Australia",
    subtitle = "Based on NASA MODIS satellite data",
    x = "Date",
    y = "Number of Fire Detections",
    color = "Avg. Confidence",
    size = "Avg. Fire Radiative Power (MW)"
  ) +
  theme(legend.position = "bottom")
```

Let's look at the distribution of fire detections by state:

```{r fire-by-state}
# Summarize fire counts by state regions
state_fires <- fire_clean %>%
  summarize(
    Total = n(),
    NSW = sum(in_nsw),
    Victoria = sum(in_victoria),
    Other = Total - NSW - Victoria
  ) %>%
  pivot_longer(cols = everything(), names_to = "Region", values_to = "Count")

# Create pie chart of fire distribution
ggplot(state_fires, aes(x = "", y = Count, fill = Region)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  scale_fill_brewer(palette = "OrRd") +
  labs(
    title = "Distribution of Fire Detections by Region",
    fill = "Region"
  ) +
  theme_void() +
  theme(legend.position = "bottom")
```

## Choropleth Map of Fire Activity

Now, let's create a choropleth map to visualize the spatial distribution of fires across Australia, focusing on fire intensity:

```{r fire-choropleth}
# Create a grid for density visualization
# This simulates a choropleth without needing actual boundaries
lon_breaks <- seq(110, 155, by = 0.5)
lat_breaks <- seq(-45, -10, by = 0.5)

fire_grid <- fire_clean %>%
  mutate(
    lon_bin = cut(longitude, breaks = lon_breaks),
    lat_bin = cut(latitude, breaks = lat_breaks)
  ) %>%
  group_by(lon_bin, lat_bin) %>%
  summarize(
    fire_count = n(),
    avg_frp = mean(frp, na.rm = TRUE),
    avg_brightness = mean(brightness, na.rm = TRUE),
    mid_lon = mean(longitude),
    mid_lat = mean(latitude),
    .groups = "drop"
  ) %>%
  filter(!is.na(lon_bin), !is.na(lat_bin))

# Display interactive choropleth map
fire_map <- leaflet(fire_grid) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~mid_lon,
    lat = ~mid_lat,
    radius = ~sqrt(fire_count) * 2,
    color = ~colorNumeric("YlOrRd", domain = avg_frp)(avg_frp),
    fillOpacity = 0.7,
    popup = ~paste(
      "Fire Count:", fire_count, "<br>",
      "Avg Fire Radiative Power:", round(avg_frp, 1), "MW<br>",
      "Avg Brightness Temp:", round(avg_brightness, 1), "K"
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("YlOrRd", domain = fire_grid$avg_frp),
    values = ~avg_frp,
    title = "Avg Fire Radiative Power (MW)",
    opacity = 0.7
  )

fire_map
```

The choropleth map reveals the concentration of fire activity in southeastern Australia, particularly in New South Wales and Victoria. The most intense fires (measured by Fire Radiative Power) occurred in forested regions along the Great Dividing Range.

## NSW Focus: The Epicenter of Crisis

New South Wales was particularly hard-hit during the 2019-2020 fire season. Let's take a closer look at the fire patterns in this state:

```{r nsw-analysis}
# Focus on NSW fires
nsw_fires <- fire_clean %>%
  filter(in_nsw)

# Create time series of NSW fire detections
nsw_daily <- nsw_fires %>%
  group_by(date) %>%
  summarize(
    detection_count = n(),
    avg_frp = mean(frp, na.rm = TRUE),
    .groups = "drop"
  )

# Plot NSW fire activity
ggplot(nsw_daily, aes(x = date, y = detection_count)) +
  geom_col(aes(fill = avg_frp), alpha = 0.8) +
  scale_fill_viridis_c(option = "inferno") +
  labs(
    title = "Fire Detection Activity in New South Wales",
    subtitle = "Number of fire detections and average intensity",
    x = "Date",
    y = "Number of Fire Detections",
    fill = "Avg. Fire\nRadiative Power (MW)"
  ) +
  theme(legend.position = "right")
```

Here is the detailed map of NSW fire activity:

```{r nsw-map}
# Create a focused map for NSW
nsw_map <- leaflet(nsw_fires) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = ~sqrt(frp)/2,
    color = ~colorNumeric("YlOrRd", domain = c(0, max(frp, na.rm = TRUE)))(frp),
    fillOpacity = 0.7,
    popup = ~paste(
      "Date:", acq_date, "<br>",
      "Time:", acq_time, "<br>",
      "Fire Radiative Power:", round(frp, 1), "MW<br>",
      "Confidence:", confidence
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = colorNumeric("YlOrRd", domain = c(0, max(nsw_fires$frp, na.rm = TRUE))),
    values = ~frp,
    title = "Fire Radiative Power (MW)",
    opacity = 0.7
  )

nsw_map
```


## Regional Fire Intensity

Let's create an additional professional visualization that shows the distribution of fire radiative power (intensity) across different regions of Australia:

```{r regional-fire-intensity}
# Create regional categories for analysis
fire_regions <- fire_clean %>%
  mutate(
    region = case_when(
      in_nsw ~ "New South Wales",
      in_victoria ~ "Victoria",
      longitude >= 114 & longitude <= 129 & latitude >= -36 & latitude <= -20 ~ "Western Australia",
      longitude >= 130 & longitude <= 141 & latitude >= -30 & latitude <= -10 ~ "Northern Territory",
      longitude >= 137 & longitude <= 154 & latitude >= -29 & latitude <= -10 ~ "Queensland",
      longitude >= 129 & longitude <= 141 & latitude >= -38 & latitude <= -30 ~ "South Australia",
      longitude >= 144 & longitude <= 149 & latitude >= -43 & latitude <= -40 ~ "Tasmania",
      TRUE ~ "Other"
    )
  ) %>%
  filter(region != "Other")

# Calculate fire statistics by region
region_stats <- fire_regions %>%
  group_by(region) %>%
  summarize(
    fire_count = n(),
    avg_frp = mean(frp, na.rm = TRUE),
    max_frp = max(frp, na.rm = TRUE),
    median_frp = median(frp, na.rm = TRUE),
    total_frp = sum(frp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_frp))

# Create a more sophisticated visualization
ggplot(region_stats, aes(x = reorder(region, total_frp), y = total_frp/1000)) +
  geom_col(aes(fill = avg_frp), width = 0.7) +
  geom_text(aes(label = paste0(round(avg_frp, 1), " MW")), 
            hjust = -0.1, vjust = 0.5, size = 3.5) +
  geom_point(aes(y = max_frp/1000), color = "red", size = 3) +
  scale_fill_viridis_c(option = "inferno", begin = 0.3, end = 0.9,
                      guide = guide_colorbar(title.position = "top")) +
  coord_flip() +
  labs(
    title = "Fire Intensity by Australian Region",
    subtitle = "Based on MODIS satellite data (Dec 2019 - Jan 2020)",
    x = NULL,
    y = "Total Fire Radiative Power (Thousands of MW)",
    fill = "Average FRP (MW)",
    caption = "Red dots indicate maximum fire radiative power detected"
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.caption = element_text(hjust = 0, face = "italic"),
    axis.text.y = element_text(face = "bold", size = 10)
  )

# Create a complementary visualization showing daily fire counts by region
fire_regions %>%
  group_by(region, date) %>%
  summarize(
    daily_fires = n(),
    daily_avg_frp = mean(frp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = date, y = daily_fires, color = region)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_point(aes(size = daily_avg_frp), alpha = 0.6) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(1, 5), name = "Avg. FRP (MW)") +
  labs(
    title = "Daily Fire Detections by Region",
    subtitle = "Size indicates average fire intensity",
    x = "Date",
    y = "Number of Fire Detections",
    color = "Region"
  ) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

This visualization provides deeper insights into the regional differences in fire intensity and temporal patterns. New South Wales and Victoria not only experienced the highest number of fires but also some of the most intense fires as measured by Fire Radiative Power (FRP). The temporal pattern shows how the crisis escalated in different regions, with NSW experiencing a sharp increase in early January 2020.

The dual approach—showing both total impact (bar chart) and temporal evolution (line chart)—offers complementary perspectives on the crisis. The use of color encoding for average intensity, combined with markers for maximum intensity, allows viewers to understand both typical and extreme fire behavior in each region.

## Narrative: Living Through the Black Summer

The 2019-2020 Australian bushfire season, often referred to as the "Black Summer," was unprecedented in its scale and intensity. I remember watching the news with increasing concern as the fires escalated. Sydney, where one of my professors at Georgetown University is from, was engulfed in smoke for weeks. The air quality became so hazardous that breathing outdoors was difficult, and the iconic blue skies turned an apocalyptic orange.

I recall a phone conversation with my professor who described the eerie feeling of seeing the sun as a dim red orb through the thick smoke. Schools were closed, outdoor events canceled, and N95 masks became essential items long before the COVID-19 pandemic made them ubiquitous. The fires didn't discriminate – they consumed national parks, farmlands, and homes alike.

Most heartbreaking was the toll on wildlife. I've always been fascinated by Australia's unique ecosystems, and the estimates of a billion animals lost left me deeply saddened. Seeing images of koalas with burnt paws being rescued or kangaroos fleeing walls of flame brought home the ecological catastrophe unfolding.

The data visualization in this analysis captures the scientific measurements of this disaster, but behind these numbers are stories of communities fighting to save their homes, wildlife carers working around the clock, and a nation coming to terms with a changing climate. As the choropleth map shows, New South Wales bore the brunt of this crisis, with fire detections clustering along the eastern forests where many rural communities are located.

What struck me most about this disaster was how clearly it demonstrated the links between climate patterns and fire behavior. The areas that experienced the most severe drought conditions and highest temperatures were precisely where the most intense fires occurred. This relationship, visible in our data analysis, underscores the warnings that climate scientists have been issuing for years.

## Conclusion

This spatial analysis of the 2019-2020 Australian bushfires has revealed several critical insights into one of the most devastating ecological disasters in recent history. The fires were heavily concentrated in southeastern Australia, with New South Wales and Victoria bearing the brunt of the crisis. Our choropleth mapping clearly demonstrates this geographical concentration, highlighting how certain ecological regions—particularly along the Great Dividing Range—experienced disproportionate fire activity.

Climate conditions played a decisive role in creating the perfect storm for these catastrophic fires. The data reveals a strong correlation between areas experiencing prolonged drought and high temperatures and subsequent severe fire activity. This relationship underscores what climate scientists have long warned: changing climate patterns can dramatically intensify fire seasons. The temporal analysis shows how quickly the situation escalated in December 2019 and January 2020, with fire detections spiking dramatically over a matter of weeks.

As Australia and the world continue to confront the realities of climate change, understanding these spatial and temporal patterns becomes increasingly vital for preparedness and mitigation strategies. The Black Summer serves as a sobering reminder of what may become more common globally as climate patterns shift. The experiences from this crisis provide valuable lessons that can inform policy and practice across fire-prone regions worldwide.

Future research could build upon this analysis in several promising directions. Incorporating longer-term climate data would provide richer historical context for understanding the uniqueness of the 2019-2020 conditions. Including vegetation and fuel load data would significantly improve fire risk modeling capabilities. Analyzing recovery patterns in affected areas through satellite imagery would help assess ecosystem resilience and inform restoration efforts. Finally, integrating social and economic impact data would help quantify the human dimension of the disaster and inform more holistic preparation and response strategies.

The choropleth mapping approach used in this analysis has demonstrated its value as a powerful tool for visualizing spatial patterns in fire activity. By combining this spatial analysis with climate data, we can develop increasingly sophisticated models to predict, prepare for, and potentially mitigate future fire disasters. The insights gained through this work contribute to our collective understanding of wildfire dynamics in a changing climate—knowledge that becomes ever more critical as we face an uncertain future.

## References

1. TidyTuesday. (2020). Australia Fires Dataset. Retrieved from https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-01-07
2. Australian Bureau of Meteorology. (2020). Climate Data Online.
3. NASA FIRMS. (2020). Fire Information for Resource Management System.
4. New South Wales Rural Fire Service. (2020). Major Incidents Feed.
5. Boer, M. M., Resco de Dios, V., & Bradstock, R. A. (2020). Unprecedented burn area of Australian mega forest fires. Nature Climate Change, 10(3), 171-172.
