---
title: "Working with Dates"
format: html
---

## Learning Goals

-   Review wrangling verbs: select, mutate, filter, arrange, summarize, group_by
-   Form conceptual understanding of code without running it
-   Practice using wrangling verbs in different ways
-   Practice dealing with dates using lubridate functions

## Additional Resources

For more information about the topics covered in this chapter, refer to the resources below:

-   [Date and Times (html)](https://r4ds.had.co.nz/dates-and-times.html) by Wickham, Çetinkaya-Rundel, & Grolemund

## 9.1 Review

### Data Science Process

Below is the visual representation of the data science process we saw earlier. Which stage are we in currently?

*We are in the wrangling stage of the data science process.*

Recall that wrangling is important. It is much of what we spend our efforts on in Data Science. There are lots of steps, hence R functions, that can go into data wrangling. But we can get far with the following 6 wrangling verbs:

| verb      | action                                    |
|-----------|-------------------------------------------|
| arrange   | arrange the rows according to some column |
| filter    | filter out or obtain a subset of the rows |
| select    | select a subset of columns                |
| mutate    | mutate or create a column                 |
| summarize | calculate a numerical summary of a column |
| group_by  | group the rows by a specified column      |

### Example 1: Single Verb

Let's start by working with some TidyTuesday data on penguins. This data includes information about penguins' flippers ("arms") and bills ("mouths" or "beaks"). Let's import this using read_csv(), a function in the tidyverse package. For the most part, this is similar to read.csv(), though read_csv() can be more efficient at importing large datasets.

```{r}
library(tidyverse)
penguins <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-28/penguins.csv')

# Check it out
head(penguins)
```

### Check Understanding

Construct a plot that allows us to examine how the relationship between body mass and bill length varies by species and sex.

```{r}
ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g, color = species, shape = sex)) +
  geom_point() +
  labs(title = "Relationship between Body Mass and Bill Length",
       x = "Bill Length (mm)",
       y = "Body Mass (g)",
       color = "Species",
       shape = "Sex") +
  theme_minimal()
```

### Check Understanding

Use the 6 wrangling verbs to address each task in the code chunk below. You can tack on \|\> head() to print out just 6 rows to keep your rendered document manageable. Most of these require just 1 verb.

```{r}
# Get data on only Adelie penguins that weigh more than 4700g
penguins |> 
  filter(species == "Adelie", body_mass_g > 4700) |>
  head()
```

```{r}
# Get data on penguin body mass only
# Show just the first 6 rows
penguins |> 
  select(body_mass_g) |>
  head()
```

```{r}
# Sort the penguins from smallest to largest body mass
# Show just the first 6 rows
penguins |> 
  arrange(body_mass_g) |>
  head()
```

```{r}
# Calculate the average body mass across all penguins
# Note: na.rm = TRUE removes the NAs from the calculation
penguins |> 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE))
```

```{r}
# Calculate the average body mass by species
penguins |> 
  group_by(species) |>
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE))
```

```{r}
# Create a new column that records body mass in kilograms, not grams
# NOTE: there are 1000 g in 1 kg
# Show just the first 6 rows
penguins |> 
  mutate(body_mass_kg = body_mass_g / 1000) |>
  head()
```

### Check Understanding

How many penguins of each species do we have? Create a viz that addresses this question.

```{r}
ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Number of Penguins by Species",
       x = "Species",
       y = "Count") +
  theme_minimal()
```

### Check Understanding

Can we use the 6 verbs to calculate exactly how many penguins in each species?

HINT: n() calculates group size.

```{r}
penguins |>
  group_by(species) |>
  summarize(count = n())
```

### count verb

The count() verb provides a handy shortcut!

```{r}
penguins |> 
  count(species)
```

### Example 2: Multiple Verbs

#### Check Understanding

Let's practice combining some verbs. For each task:

-   Translate the prompt into our 6 verbs. That is, think before you type.
-   Build your code line by line. It's important to understand what's being piped into each function!
-   Ask what you can rearrange and still get the same result.
-   Read your final code like a paragraph / a conversation. Would another person be able to follow your logic?

```{r}
# Sort Gentoo penguins from biggest to smallest with respect to their 
# bill length in cm (there are 10 mm in a cm)
penguins |>
  filter(species == "Gentoo") |>
  mutate(bill_length_cm = bill_length_mm / 10) |>
  arrange(desc(bill_length_cm))
```

```{r}
# Sort the species from smallest to biggest with respect to their 
# average bill length in cm
penguins |>
  group_by(species) |>
  summarize(avg_bill_length_cm = mean(bill_length_mm, na.rm = TRUE) / 10) |>
  arrange(avg_bill_length_cm)
```

### Example 3: Interpret Code

Let's practice reading and making sense of somebody else's code. What do you think this produces?

-   How many columns? Rows?
-   What are the column names?
-   What's represented in each row?

Once you've thought about it, put the code inside a chunk and run it!

```{r}
penguins |> 
  filter(species == "Chinstrap") |> 
  group_by(sex) |> 
  summarize(min = min(body_mass_g, na.rm = TRUE), 
            max = max(body_mass_g, na.rm = TRUE)) |> 
  mutate(range = max - min)
```

## 9.2 Exercises Part 1: Same Verbs, New Tricks

### Exercise 1: More Filtering

Recall the "logical comparison operators" we can use to filter() our data:

| symbol           | meaning                   |
|------------------|---------------------------|
| ==               | equal to                  |
| !=               | not equal to              |
| \>               | greater than              |
| \>=              | greater than or equal to  |
| \<               | less than                 |
| \<=              | less than or equal to     |
| %in% c(***,*** ) | a list of multiple values |

#### Part a

```{r}
# Create a dataset with just Adelie and Chinstrap using %in%
# Pipe this into `count(species)` to confirm that you only have these 2 species
penguins |> 
  filter(species %in% c("Adelie", "Chinstrap")) |> 
  count(species)
```

```{r}
# Create a dataset with just Adelie and Chinstrap using !=
# Pipe this into `count(species)` to confirm that you only have these 2 species
penguins |> 
  filter(species != "Gentoo") |> 
  count(species)
```

#### Part b

Notice that some of our penguins have missing (NA) data on some values:

```{r}
head(penguins)
```

##### Handling NA Values

There are many ways to handle missing data. The right approach depends upon your research goals. A general rule is: Only get rid of observations with missing data if they're missing data on variables you need for the specific task at hand!

###### Example 1

Suppose our research focus is just on body_mass_g. Two penguins are missing this info:

```{r}
# NOTE the use of is.na()
penguins |> 
  summarize(sum(is.na(body_mass_g)))
```

Let's define a new dataset that removes these penguins:

```{r}
# NOTE the use of is.na()
penguins_w_body_mass <- penguins |> 
  filter(!is.na(body_mass_g))

# Compare the number of penguins in this vs the original data
nrow(penguins_w_body_mass)
```

```{r}
nrow(penguins)
```

Note that some penguins in penguins_w_body_mass are missing info on sex, but we don't care since that's not related to our research question:

```{r}
penguins_w_body_mass |> 
  summarize(sum(is.na(sex)))
```

###### Example 2

In the very rare case that we need complete information on every variable for the specific task at hand, we can use na.omit() to get rid of any penguin that's missing info on any variable:

```{r}
penguins_complete <- penguins |> 
  na.omit()

# How many penguins did this eliminate?
nrow(penguins_complete)
```

```{r}
nrow(penguins)
```

#### Part c

Explain why we should only use na.omit() in extreme circumstances.

*We should only use na.omit() in extreme circumstances because it removes entire observations (rows) that have ANY missing values in ANY columns, which can drastically reduce your dataset size and potentially introduce bias. When we remove data, we're making assumptions about the missing values that may not be valid. It's better to only remove observations with missing values in specific columns that are directly related to the analysis at hand, rather than eliminating all incomplete observations.*

### Exercise 2: More Selecting

Being able to select() only certain columns can help simplify our data. This is especially important when we're working with lots of columns (which we haven't done yet). It can also get tedious to type out every column of interest. Here are some shortcuts:

-   "-" removes a given variable and keeps all others (e.g. select(-island))
-   starts_with("***"), ends_with("***"), or contains("\_\_\_") selects only the columns that either start with, end with, or simply contain the given string of characters

Use these shortcuts to create the following datasets.

```{r}
# First: recall the variable names
names(penguins)
```

```{r}
# Use a shortcut to keep everything but the year and island variables
penguins |>
  select(-year, -island) |>
  head()
```

```{r}
# Use a shortcut to keep only species and the penguin characteristics measured in mm
penguins |>
  select(species, ends_with("_mm")) |>
  head()
```

```{r}
# Use a shortcut to keep only species and bill-related measurements
penguins |>
  select(species, starts_with("bill")) |>
  head()
```

```{r}
# Use a shortcut to keep only species and the length-related characteristics
penguins |>
  select(species, contains("length")) |>
  head()
```

### Exercise 3: Arranging, Counting, & Grouping by Multiple Variables

We've done examples where we need to filter() by more than one variable, or select() more than one variable. Use your intuition for how we can arrange(), count(), and group_by() more than one variable.

```{r}
# Change this code to sort the penguins by species, and then island name
# NOTE: The first row should be an Adelie penguin living on Biscoe island
penguins |> 
  arrange(species, island) |>
  head()
```

```{r}
# Change this code to count the number of male/female penguins observed for each species
penguins |> 
  count(species, sex)
```

```{r}
# Change this code to calculate the average body mass by species and sex
penguins |> 
  group_by(species, sex) |> 
  summarize(mean = mean(body_mass_g, na.rm = TRUE))
```

### Exercise 4: Dates

Before some wrangling practice, let's explore another important concept: working with or mutating date variables. Dates are a whole special object type or class in R that automatically respect the order of time.

```{r}
# Get today's date
today <- Sys.Date()
today
```

```{r}
# Check out the class of this object
class(today)
```

The lubridate package inside tidyverse contains functions that can extract various information from dates. Let's learn about some of the most common functions by applying them to today. For each, make a comment on what the function does

```{r}
library(lubridate)

year(today)  # Extracts the year from the date
```

```{r}
# What do these lines produce / what's their difference?
month(today)  # Extracts the month as a number (1-12)

month(today, label = TRUE)  # Extracts the month as a labeled factor (Jan-Dec)
```

```{r}
# What does this number mean?
week(today)  # Returns the week number of the year (1-53)
```

```{r}
# What do these lines produce / what's their difference?
mday(today)  # Returns the day of the month (1-31)

yday(today)  # Returns the day of the year (1-366)
```

```{r}
# What do these lines produce / what's their difference?
wday(today)  # Returns the day of the week as a number (1=Sunday, 7=Saturday)

wday(today, label = TRUE)  # Returns the day of the week as a labeled factor
```

```{r}
# What do the results of these 2 lines tell us?
today >= ymd("2024-02-14")  # Checks if today is on or after Feb 14, 2024

today < ymd("2024-02-14")  # Checks if today is before Feb 14, 2024
```

## 9.3 Exercises Part 2: Application

The remaining exercises are similar to some of those on the homework. Hence, the solutions are not provided. Let's apply these ideas to the daily Birthdays dataset in the mosaic package.

```{r}
library(mosaic)
data("Birthdays")
head(Birthdays)
```

Birthdays gives the number of births recorded on each day of the year in each state from 1969 to 1988. We can use our wrangling skills to understand some drivers of daily births. Putting these all together can be challenging! Remember the following ways to make tasks more manageable:

-   Translate the prompt into our 6 verbs (and count()). That is, think before you type.
-   Build your code line by line. It's important to understand what's being piped into each function!

### Exercise 5: Warming up

```{r}
# How many days of data do we have for each state?
Birthdays |>
  count(state)
```

```{r}
# How many total births were there in this time period?
Birthdays |>
  summarize(total_births = sum(births))
```

```{r}
# How many total births were there per state in this time period, sorted from low to high?
Birthdays |>
  group_by(state) |>
  summarize(total_births = sum(births)) |>
  arrange(total_births)
```

### Exercise 6: Homework Reprise

Create a new dataset named daily_births that includes the total number of births per day (across all states) and the corresponding day of the week, eg, Mon. NOTE: Name the column with total births so that it's easier to wrangle and plot.

```{r}
daily_births <- Birthdays |>
  group_by(date, wday) |>
  summarize(total_births = sum(births), .groups = 'drop')

head(daily_births)
```

Using this data, construct a plot of births over time, indicating the day of week.

```{r}
ggplot(daily_births, aes(x = date, y = total_births, color = wday)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(title = "Daily Births in the US (1969-1988)",
       x = "Date",
       y = "Total Births",
       color = "Day of Week") +
  theme_minimal()
```

### Exercise 7: Wrangle & Plot

For each prompt below, you can decide whether you want to: (1) wrangle and store data, then plot; or (2) wrangle data and pipe directly into ggplot. For example:

```{r}
penguins |> 
  filter(species != "Gentoo") |> 
  ggplot(aes(y = bill_length_mm, x = bill_depth_mm, color = species)) + 
    geom_point()
```

#### Part a

Calculate the total number of births in each month and year, eg, Jan 1969, Feb 1969, …. Label month by names not numbers, eg, Jan not 1. Then, plot the births by month and comment on what you learn.

```{r}
# Calculate births by month and year
monthly_births <- Birthdays |>
  mutate(month_name = month(date, label = TRUE),
         year_month = paste(year, month_name)) |>
  group_by(year, month_name) |>
  summarize(total_births = sum(births), .groups = 'drop')

# Plot the data
ggplot(monthly_births, aes(x = month_name, y = total_births, group = year, color = as.factor(year))) +
  geom_line() +
  labs(title = "Monthly Births (1969-1988)",
       x = "Month",
       y = "Total Births",
       color = "Year") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide the legend as there are too many years
```

#### Part b

In 1988, calculate the total number of births per week in each state. Get rid of week "53", which isn't a complete week! Then, make a line plot of births by week for each state and comment on what you learn. For example, do you notice any seasonal trends? Are these the same in every state? Any outliers?

```{r}
# Filter for 1988 data, calculate weekly totals by state
weekly_births_1988 <- Birthdays |>
  filter(year == 1988) |>
  mutate(week_num = week(date)) |>
  filter(week_num != 53) |>  # Remove week 53
  group_by(state, week_num) |>
  summarize(weekly_births = sum(births), .groups = 'drop')

# Plot for some selected states (too many to show all)
selected_states <- c("CA", "NY", "TX", "FL", "IL")

weekly_births_1988 |>
  filter(state %in% selected_states) |>
  ggplot(aes(x = week_num, y = weekly_births, color = state)) +
  geom_line() +
  labs(title = "Weekly Births in 1988 by State",
       x = "Week Number",
       y = "Total Births",
       color = "State") +
  theme_minimal()
```

#### Part c

Repeat the above for just Minnesota (MN) and Louisiana (LA). MN has one of the coldest climates and LA has one of the warmest. How do their seasonal trends compare? Do you think these trends are similar in other colder and warmer states? Try it!

```{r}
# Filter for Minnesota and Louisiana in 1988
mn_la_weekly <- Birthdays |>
  filter(year == 1988, state %in% c("MN", "LA")) |>
  mutate(week_num = week(date)) |>
  filter(week_num != 53) |>  # Remove week 53
  group_by(state, week_num) |>
  summarize(weekly_births = sum(births), .groups = 'drop')

# Plot comparing MN and LA
ggplot(mn_la_weekly, aes(x = week_num, y = weekly_births, color = state)) +
  geom_line() +
  labs(title = "Weekly Births in 1988: Minnesota vs Louisiana",
       x = "Week Number",
       y = "Total Births",
       color = "State") +
  theme_minimal()
```

### Exercise 8: More Practice

#### Part a

Create a dataset with only births in Massachusetts (MA) in 1979 and sort the days from those with the most births to those with the fewest.

```{r}
ma_1979 <- Birthdays |>
  filter(state == "MA", year == 1979) |>
  arrange(desc(births))

head(ma_1979)
```

#### Part b

Make a table showing the five states with the most births between September 9, 1979 and September 12, 1979, including the 9th and 12th. Arrange the table in descending order of births.

```{r}
sept_1979_births <- Birthdays |>
  filter(date >= ymd("1979-09-09"), date <= ymd("1979-09-12")) |>
  group_by(state) |>
  summarize(total_births = sum(births)) |>
  arrange(desc(total_births)) |>
  head(5)

sept_1979_births
```
